Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2016-02-04 pldi-2016/paper.tex
Further edits - sections 3 and 4

\item We give a formal model (\S\ref{sec:formal}) and use it to prove
  \emph{relativized type safety} for the type providers (\S\ref{sec:safety}).
\end{itemize}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.80,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
\end{center}
\vspace{-0.5em}
\caption{Important aspects of the preferred shape relation}
\label{fig:subtyping-diagram}
\vspace{-0.5em}
\end{figure}

@@

We distinguish between \emph{non-nullable shapes} that always have a valid value (written as 
$\hat{\sigma}$) and \emph{nullable shapes} that encompass missing and \kvd{null} values 
(written as $\sigma$). We write $\nu$ for record names and record field names.

\noindent
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n,\; \rho_i  \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{nullable}\langl \hat{\sigma} \rangl \lsep [\sigma] \lsep \kvd{any} \lsep \kvd{null}  \lsep ~\bot~
\end{array}
\end{equation*}

@@

Any non-nullable shape $\hat{\sigma}$ can be wrapped as $\kvd{nullable}\langl\hat{\sigma}\rangl$ to 
explicitly permit the \kvd{null} value. Type providers map \kvd{nullable} shapes to the F\# option
type. A collection $[\sigma]$ is also nullable and \kvd{null} values are treated as empty 
collections. The shape $\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded 
notation) and $\bot$ is the bottom shape. The \kvd{any} shape is the top shape, but we revisit it 
later by adding labels for statically known alternative shapes (\S\ref{sec:inference-vars}) as 
discussed earlier (\S\ref{sec:providers-xml}).

During inference we use row-variables $\rho_i$ \cite{rows-cardelli} in record shapes to represent the flexibility
arising from records in samples. For example, when a 
record $\ident{Point}\,\{ \ident{x} \mapsto \num{3} \}$ occurs in a sample,
it may be combined through inference with other
samples such as $\ident{Point}\,\{\, \ident{x}\mapsto\num{3}, \ident{y}\mapsto\num{4} \}$ that contain more fields. The
overall shape inferred must account for the fact that any extra fields are optional,
giving an inferred shape $\ident{Point}\,\{ \ident{x}\!:\!\ident{int},~ \ident{y}\!:\!\kvd{option}\langl\ident{int}\rangl\}$.
The process of solving $\rho_i$ variables is discussed in \S\ref{sec:formal-inferval}.

@@

Figure~\ref{fig:subtyping-diagram} provides an intuition about the preference between 
shapes. The upper part shows non-nullable shapes (with records and primitives) and the lower part 
shows nullable shapes with \kvd{null}, collections and nullable shapes. In the diagram, we 
abbreviate $\kvd{nullable}\langl\sigma\rangl$ as $\sigma?$ and we omit links between the two parts;
a shape $\hat{\sigma}$ is preferred over $\kvd{nullable}\langl\hat{\sigma}\rangl$.

\begin{definition}
For ground $\sigma_1$, $\sigma_2$ (i.e. without $\rho_i$ variables), we write $\sigma_1 \sqsupseteq \sigma_2$ to denote that $\sigma_2$ is preferred over $\sigma_1$. 
The shape preference relation is defined as a transitive reflexive closure of the following rules:

\noindent
\begin{align}
  \label{eq:sub-prim}\tag{P1}
  \ident{float}\,&\sqsupseteq\,\ident{int}&\\[-0.2em]
  \label{eq:sub-null}\tag{P2}
  \sigma &\sqsupseteq \kvd{null}  &(\textnormal{for}~\sigma \neq \hat{\sigma})  \\[-0.2em]
  \label{eq:sub-opt}\tag{P3}
  \kvd{nullable}\langl\hat{\sigma}\rangl &\sqsupseteq \hat{\sigma}  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
  \label{eq:sub-opt-cov}\tag{P4}
  \kvd{nullable}\langl\hat{\sigma_1}\rangl &\sqsupseteq 
    \kvd{nullable}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} \sqsupseteq \hat{\sigma_2})\\[-0.2em]
  \label{eq:sub-col}\tag{P5}
  [\sigma_1] &\sqsupseteq [\sigma_2]  &(\textnormal{if}~\sigma_1 \sqsupseteq \sigma_2) \\[-0.2em]
  \label{eq:sub-bot}\tag{P6}
  \sigma &\sqsupseteq \bot  &(\textnormal{for all}~\sigma)\\[-0.2em]
  \label{eq:sub-var-top}\tag{P7}
  \kvd{any} &\sqsupseteq \sigma 
\end{align}
\vspace{-2em}

\noindent
\begin{align}
\label{eq:sub-record1}\tag{R1}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsupseteq \\
 \quad\nu~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
\end{array} \qquad ~~~~(\textnormal{if}~\sigma_i \sqsupseteq \sigma_i')
\end{align}

@@

\begin{align}
\label{eq:sub-record2}\tag{R2}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsupseteq \\
 \quad\nu~\{ \nu_1\!:\!\sigma_1, .., .., \nu_m\!:\!\sigma_m \}
\end{array} \quad~~\, (\textnormal{when}~m \geq n)
\end{align}

@@

\begin{figure*}[t]
\noindent
\begin{equation*}
\begin{array}{rclcl}
 \tytag &\narrow{=}& \ident{collection}  &\narrow{\lsep}& \ident{number} \\
        &\narrow{\lsep}& \ident{nullable} &\narrow{\lsep}& \ident{string}  \\
        &\narrow{\lsep}& \nu ~\lsep \ident{any} &\narrow{\lsep}& \ident{bool}
\end{array}
%
\quad\;\;\;
%
\begin{array}{rcl}
 \tytagof(\ident{string}) &\narrow{=}& \ident{string}\\
 \tytagof(\ident{bool}) &\narrow{=}& \ident{bool}\\
 \tytagof(\ident{int}) &\narrow{=}& \ident{number}\\
 \tytagof(\ident{float}) &\narrow{=}& \ident{number}\\
\end{array}
%
\;\;
%
\begin{array}{rcl}
 \tytagof(\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) &\narrow{=}& \ident{any}\\
 \tytagof(\nu\; \{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \nu \\
 \tytagof(\kvd{nullable}\langl\hat{\sigma}\rangl) &\narrow{=}& \ident{nullable}\\
 \tytagof([\sigma]) &\narrow{=}& \ident{collection}\\
\end{array}
\end{equation*}

\vspace{-0.5em}
\begin{equation*}
\quad\qquad\qquad
\begin{array}{rcll}
 \addopt{\hat{\sigma}} &\narrow{=}& \kvd{nullable}\langl\hat{\sigma}\rangl &(\textnormal{non-nullable shapes})\\
 \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\qquad\qquad
\begin{array}{rcll}
 \dropopt{\kvd{nullable}\langl\hat{\sigma}\rangl} &\narrow{=}& \hat{\sigma} &(\textnormal{nullable shape})\\
 \dropopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\end{equation*}

{\color{red} TO BE DELETED}

\vspace{-0.5em}
\begin{equation*}
% option types
\color{red}
\inference[(nullable)\;]
  {\hat{\sigma_1} \tsep \sigma_2 \vdash \sigma}
  {\kvd{nullable}\langl\hat{\sigma_1}\rangl \tsep \sigma_2 \vdash \kvd{nullable}\langl\dropopt{\sigma}\rangl}
\qquad
\inference[(any)\;]
  { \tytagof(\sigma_1) \neq \tytagof(\sigma_2) & \tytagof(\sigma_1) \neq \ident{nullable} \neq \tytagof{\sigma_2}}
  { \sigma_1 \tsep \sigma_2 \vdash \kvd{any} }  
\end{equation*}
\vspace{-2em}

\begin{equation*}
\color{red}
\hspace{4em}
\inference[(record)\;]
  { \forall i\in\{ 1 .. n \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
  { \begin{array}{l}
    \nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \} \tsep
    \nu \; \{ \nu'_1 \!:\! \sigma'_1, \; \ldots \;, \nu'_m \!:\! \sigma'_m \} \vdash
    \nu \; \{ \nu_1 \!:\! \sigma''_1, \; \ldots \; , \nu_k \!:\! \sigma''_n \}
    \end{array} }
\end{equation*}
\vspace{-2em}

% all rules are symmetric and reflexive
\begin{equation*}
\color{red}
\hspace{2em}
\inference[(list)\;]
  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
  {[\sigma_1] \tsep [\sigma_2] \vdash [\sigma]}
\quad
\inference[(sym)\;]
  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
  {\sigma_2 \tsep \sigma_1 \vdash \sigma}
%
\quad
\begin{array}{l}
 \textnormal{\footnotesize{(refl)}}\;\; \sigma \tsep \sigma \vdash \sigma\\[0.6em]
 % bottom type
 \textnormal{\footnotesize{(bot)}}\;\; \bot \tsep \sigma \vdash \sigma
\end{array}
%
\quad
% null and primitives
\begin{array}{l}
 \textnormal{\footnotesize{(null)}}\;\; \sigma \tsep \kvd{null} \vdash \addopt{\sigma} \quad (\sigma \neq \bot)\\[0.6em]
 \textnormal{\footnotesize{(prim)}}\;\; \kvd{int} \tsep \kvd{float} \vdash \kvd{float}
\end{array}
\end{equation*}

\begin{equation*}
\begin{array}{rcl}
\ident{lub}(\sigma, \sigma) &=& \sigma \\
\ident{lub}([\sigma_1], [\sigma_2]) &=& [\ident{lub}(\sigma_1, \sigma_2)]  \\
\ident{lub}(\bot, \sigma) = \ident{lub}(\sigma, \bot) &=& \sigma \\
\ident{lub}(\kvd{null}, \sigma) = \ident{lub}(\sigma, \kvd{null}) &=& \addopt{\sigma} \\
\ident{lub}(\kvd{any}, \sigma) = \ident{lub}(\sigma, \kvd{any}) &=& \kvd{any} \\
\ident{lub}(\kvd{float}, \kvd{int}) = \ident{lub}(\kvd{int}, \kvd{float}) &=& \kvd{float} \\
\ident{lub}(\sigma_2, \kvd{nullable}\langl \hat{\sigma_1} \rangl) = \ident{lub}(\kvd{nullable}\langl \hat{\sigma_1} \rangl, \sigma_2) &=& \addopt{\ident{lub}(\hat{\sigma_1}, \sigma_2)} \\
\ident{lub}(\nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \}, 
  \nu \; \{ \nu_1 \!:\! \sigma'_1, \; \ldots \;, \nu_n \!:\! \sigma'_n \}) &=& 
  \nu \; \{ \nu_1 \!:\! \ident{lub}(\sigma_1, \sigma'_1), \; \ldots \; , \nu_n \!:\! \ident{lub}(\sigma_n, \sigma'_n) \}  \\[0.75em]
\textnormal{Otherwise}~~~~ \ident{lub}(\sigma_1, \sigma_2) &=& \kvd{any} \\

\end{array}
\end{equation*}

@@

Given two ground shapes, the \emph{common preferred shape} relation finds a least upper bound
(Lemma~\ref{thm:lub}). It prefers records, which is important for usability as discussed
earlier (\S\ref{sec:providers-xml}).

\begin{definition}
A \emph{common preferred shape} of two ground shapes $\sigma_1$ and $\sigma_2$ is a shape $\sigma$, written 
$\sigma_1 \tsep \sigma_2 \vdash \sigma$, by the rules in 
Figure~\ref{fig:subtyping-cst}.
\end{definition}

\noindent
We define shape \emph{tags} to identify shapes that have a common preferred shape which is not the
top shape. When finding a common shape of two records (\emph{record}), the shape of fields becomes the common preferred shape of their 
respective shapes. The field labels of two record types can always be made to match,
through the shape inference from sample (\S\ref{sec:formal-inferval}),
\emph{before} finding a common preferred shape using $\triangledown$.

We can find a common shape of two different numbers (\emph{prim}); for two collections,
we combine their elements (\emph{list}). When one shape is nullable (\emph{nullable}), we find
the common non-nullable shape and wrap it in \kvd{nullable}; if one of the shapes is \kvd{null}, 
we make the other one nullable (\emph{null}). Finally, if the two shapes do not have matching tags 
(and they are not nullable), we have to infer the \kvd{any} shape as there is no better alternative. 
The remaining rules for reflexivity, symmetry and the bottom shape are standard.

\paragraph{Properties.}
The partially ordered set of ground shapes has a unique least upper bound.
The $\triangledown$ relation defines a function (Lemma~\ref{thm:func}) that finds 
the least upper bound (Lemma~\ref{thm:lub}).

\begin{lemma}[Preferred shape function]
\label{thm:func}
For all ground $\sigma_1$ and $\sigma_2$ there exists exactly one $\sigma$ such that 
$\sigma_1 \tsep \sigma_2 \vdash \sigma$. 
\end{lemma}
\begin{proof}
The pre-conditions of rules in Figure~\ref{fig:subtyping-cst} are disjoint, with the exception 
of (\emph{sym}) and (\emph{refl}), but applying those does not affect the result. Reflexivity is
preserved recursively and symmetry does not enable different derivations.
\end{proof}

\begin{lemma}[Least upper bound]
\label{thm:lub}
For ground $\sigma_1$ and $\sigma_2$, if $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ then $\sigma$ is a least upper bound by $\sqsupseteq$.
\end{lemma}
\begin{proof}
By induction over $\vdash$. Note that the algorithm only infers the top shape \kvd{any} when for non-nullable
shapes of distinct tags and so no better preferred shape exists.
\end{proof}

@@

Figure~\ref{fig:shape-inference} defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a collection of sample 
data $d_1, \ldots, d_n$ into a shape $\sigma$. Before applying $S$, we assume each record in each $d_i$ is marked with a fresh row inference
variable $\rho_i$.  
We then choose a ground, minimal substitution $\theta$ for row variables.
Because $\rho_i$ variables represent potentially missing fields,
the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} is applied to all types
in the vector. 
This is sufficient to equate the record field labels and satisfy the side conditions in rule (record) when multiple record samples are merged.
In practice, $\theta$ is found via row variable unification \cite{rows-remy}. We omit the details here.
No $\rho_i$ variables remain after inference as the substitution chosen is ground.

@@

\begin{figure*}
  \noindent
  \begin{equation*}
  \textnormal{\footnotesize{(top-1)}}\;\;
  \inference
    {\exists i . \tytagof(\sigma_i) = \tytagof(\dropopt{\sigma}) & \sigma \tsep \sigma_i \vdash \sigma_i' & \tytagof(\sigma)\neq\ident{any}}
    {\sigma \tsep \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \vdash 
      \kvd{any}\langl\sigma_1, \ldots, \dropopt{\sigma_i'}, \ldots, \sigma_n\rangl}
  \;\;
  \inference
    {\nexists i . \tytagof(\sigma_i) = \tytagof(\dropopt{\sigma}) & \tytagof(\sigma)\neq\ident{any}}
    {\sigma \tsep \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \vdash 
      \kvd{any}\langl\sigma_1, \ldots, \sigma_n, \dropopt{\sigma}\rangl}
  \end{equation*}
  \vspace{-2em}

  % two union types
  \begin{equation*}
  \begin{array}{l}
  \inference[(top-2)\;]
    { (\tytagof(\sigma_i) = \tytagof(\sigma'_j)) \Leftrightarrow (i = j) \wedge (i \leq k)\\
        \forall i\in\{ 1 .. k \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
    { \begin{array}{l}
      \kvd{any}\langl \sigma_1, \ldots, \sigma_k,  \ldots, \sigma_n\rangl \tsep 
      \kvd{any}\langl \sigma'_1, \ldots, \sigma'_k, \ldots, \sigma'_m\rangl \vdash\\
      \kvd{any}\langl \sigma''_1, \ldots, \sigma''_k, \sigma_{k+1}, \ldots, \sigma_{n}, \sigma'_{k+1}, \ldots, \sigma'_{m}\rangl
      \end{array} }
  \end{array}    
  \begin{array}{l}
  \inference[(top-3)\;]
    {(\forall i\in\{1,2\}) & \tytagof(\sigma_1) \neq \tytagof(\sigma_2) \\ 
     \tytagof(\sigma_i) \neq \ident{any} & \nexists\sigma_i' . (\sigma_i = \kvd{option}\langl\sigma_i'\rangl)  }
    {\sigma_1 \tsep \sigma_2 \vdash \kvd{any}\langle\dropopt{\sigma_1}, \dropopt{\sigma_2}\rangle}
  \end{array}  
  \end{equation*}
  \caption{Extending the common preferred shape relation for labelled top shapes}
  \label{fig:subtyping-cst-var}
\end{figure*}

@@

\section{Relativized type safety}
\label{sec:safety}

>>>>

\item We give a formal model (\S\ref{sec:formal}) and use it to prove
  \emph{relative type safety} for the type providers (\S\ref{sec:safety}).
\end{itemize}

@@

@@

We distinguish between \emph{non-nullable shapes} that always have a valid value (written as 
$\hat{\sigma}$) and \emph{nullable shapes} that encompass missing and \kvd{null} values 
(written as $\sigma$). We write $\nu$ for record names and record field names. 
%
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n,\; \rho_i  \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{nullable}\langl \hat{\sigma} \rangl \lsep [\sigma] \lsep \kvd{any} \lsep \kvd{null}  \lsep ~\bot~
\end{array}
\end{equation*}

@@

Any non-nullable shape $\hat{\sigma}$ can be wrapped as $\kvd{nullable}\langl\hat{\sigma}\rangl$ to 
explicitly permit the \kvd{null} value. Type providers map \kvd{nullable} shapes to the F\# option
type. A collection $[\sigma]$ is also nullable and \kvd{null} values are treated as empty 
collections. This is motivated by the fact that a \kvd{null} collection is usually
handled as an empty collection by client code. However there is a range of design alternatives (make collections
non-nullable or treat \kvd{null} \ident{string} as an empty string).

The shape $\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded 
notation) and $\bot$ is the bottom shape. The \kvd{any} shape is the top shape, but we later add
labels for statically known alternative shapes (\S\ref{sec:inference-vars}) as 
discussed earlier (\S\ref{sec:providers-xml}).

During inference we use row-variables $\rho_i$ \cite{rows-cardelli} in record shapes to represent 
the flexibility arising from records in samples. For example, when a 
record $\ident{Point}\,\{ \ident{x} \mapsto \num{3} \}$ occurs in a sample,
it may be combined with $\ident{Point}\,\{\, \ident{x}\mapsto\num{3}, \ident{y}\mapsto\num{4} \}$ that contain more fields. The
overall shape inferred must account for the fact that any extra fields are optional,
giving an inferred shape $\ident{Point}\,\{ \ident{x}\!:\!\ident{int},~ \ident{y}\!:\!\kvd{nullable}\langl\ident{int}\rangl\}$.

\begin{figure}
\begin{center}
\includegraphics[scale=0.80,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
\end{center}
\vspace{-0.5em}
\caption{Important aspects of the preferred shape relation}
\label{fig:subtyping-diagram}
\end{figure}

@@

Figure~\ref{fig:subtyping-diagram} provides an intuition about the preference between 
shapes. The lower part shows non-nullable shapes (with records and primitives) and the upper part 
shows nullable shapes with \kvd{null}, collections and nullable shapes. In the diagram, we 
abbreviate $\kvd{nullable}\langl\sigma\rangl$ as $\sigma?$ and we omit links between the two parts;
a shape $\hat{\sigma}$ is preferred over $\kvd{nullable}\langl\hat{\sigma}\rangl$.

\begin{definition}
For ground $\sigma_1$, $\sigma_2$ (i.e. without $\rho_i$ variables), we write $\sigma_1 \sqsubseteq \sigma_2$ to denote that $\sigma_1$ is preferred over $\sigma_2$. 
The shape preference relation is defined as a transitive reflexive closure of the following rules:

\noindent
\begin{align}
  \label{eq:sub-prim}\tag{P1}
  \ident{int}\,&\sqsubseteq\,\ident{float}&\\[-0.2em]
  \label{eq:sub-null}\tag{P2}
  \kvd{null} &\sqsubseteq \sigma  &(\textnormal{for}~\sigma \neq \hat{\sigma})  \\[-0.2em]
  \label{eq:sub-opt}\tag{P3}
  \hat{\sigma} &\sqsubseteq \kvd{nullable}\langl\hat{\sigma}\rangl  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
  \label{eq:sub-opt-cov}\tag{P4}
  \kvd{nullable}\langl\hat{\sigma_1}\rangl &\sqsubseteq 
    \kvd{nullable}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} \sqsubseteq \hat{\sigma_2})\\[-0.2em]
  \label{eq:sub-col}\tag{P5}
  [\sigma_1] &\sqsubseteq [\sigma_2]  &(\textnormal{if}~\sigma_1 \sqsubseteq \sigma_2) \\[-0.2em]
  \label{eq:sub-bot}\tag{P6}
  \bot &\sqsubseteq \sigma  &(\textnormal{for all}~\sigma)\\[-0.2em]
  \label{eq:sub-var-top}\tag{P7}
  \sigma &\sqsubseteq \kvd{any}
\end{align}
\vspace{-2em}

\noindent
\begin{align}
\label{eq:sub-record1}\tag{R1}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsubseteq \\
 \quad\nu~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
\end{array} \qquad ~~~~(\textnormal{if}~\sigma_i \sqsubseteq \sigma_i')
\end{align}

@@

\begin{align}
\label{eq:sub-record2}\tag{R2}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsubseteq \\
 \quad\nu~\{ \nu_1\!:\!\sigma_1, .., .., \nu_m\!:\!\sigma_m \}
\end{array} \quad~~\, (\textnormal{when}~m \leq n)
\end{align}

@@

\begin{figure*}[t]
\noindent
\begin{equation*}
\begin{array}{rcll}
\ident{csh}(\sigma, \sigma) &=& \sigma & (\emph{eq})\\
\ident{csh}([\sigma_1], [\sigma_2]) &=& [\ident{csh}(\sigma_1, \sigma_2)] & (\emph{list}) \\
\ident{csh}(\bot, \sigma) = \ident{csh}(\sigma, \bot) &=& \sigma & (\emph{bot})\\
\ident{csh}(\kvd{null}, \sigma) = \ident{csh}(\sigma, \kvd{null}) &=& \addopt{\sigma}& (\emph{null}) \\
\ident{csh}(\kvd{any}, \sigma) = \ident{csh}(\sigma, \kvd{any}) &=& \kvd{any} & (\emph{top})\\
\ident{csh}(\kvd{float}, \kvd{int}) = \ident{csh}(\kvd{int}, \kvd{float}) &=& \kvd{float} & (\emph{num})\\
\ident{csh}(\sigma_2, \kvd{nullable}\langl \hat{\sigma_1} \rangl) = \ident{csh}(\kvd{nullable}\langl \hat{\sigma_1} \rangl, \sigma_2) &=& \addopt{\ident{csh}(\hat{\sigma_1}, \sigma_2)} & (\emph{opt})\\
\ident{csh}(\nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \}, 
  \nu \; \{ \nu_1 \!:\! \sigma'_1, \; \ldots \;, \nu_n \!:\! \sigma'_n \}) &=& 
  \nu \; \{ \nu_1 \!:\! \ident{csh}(\sigma_1, \sigma'_1), \; \ldots \; , \nu_n \!:\! \ident{csh}(\sigma_n, \sigma'_n) \} & (\emph{recd}) \\
\ident{csh}(\sigma_1, \sigma_2) &=& \kvd{any} & (\emph{any})\\
\end{array}
\end{equation*}
%
\begin{equation*}
\quad\qquad
\begin{array}{rcll}
 \addopt{\hat{\sigma}} &\narrow{=}& \kvd{nullable}\langl\hat{\sigma}\rangl &(\textnormal{non-nullable shapes})\\
 \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\qquad\qquad
\begin{array}{rcll}
 \dropopt{\kvd{nullable}\langl\hat{\sigma}\rangl} &\narrow{=}& \hat{\sigma} &(\textnormal{nullable shape})\\
 \dropopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\end{equation*}

@@

Given two ground shapes, the \emph{common preferred shape} is the least upper bound of the
shape with respect to the preferred shape relation. The least upper bound prefers records, 
which is important for usability as discussed earlier (\S\ref{sec:providers-xml}).

\begin{definition}
A \emph{common preferred shape} of two ground shapes $\sigma_1$ and $\sigma_2$ is a shape 
$\ident{csh}(\sigma_1, \sigma_2)$ obtained according to Figure~\ref{fig:subtyping-cst}. 
The rules are matched from top to bottom.
\end{definition}

\noindent
The fact that the rules of \ident{csh} are matched from top to bottom resolves the ambiguity 
between certain rules. Most importantly (\emph{any}) is used only as the last resort.
When finding a common shape of two records (\emph{recd}) we find common preferred shapes of their 
respective fields. The field labels of two record types can always be made to match, through the 
shape inference from sample (\S\ref{sec:formal-inferval}), \emph{before} finding a common preferred 
shape using $\ident{csh}$.

We can find a common shape of two different numbers (\emph{num}); for two collections,
we combine their elements (\emph{list}). When one shape is nullable (\emph{opt}), we find
the common non-nullable shape and ensure the result is nullable using $\addopt{-}$, which
is also applied when one of the shapes is \kvd{null} (\emph{null}).

The \ident{csh} function finds the unique least upper bound of the partially ordered 
set of ground shapes (Lemma~\ref{thm:lub}).

\begin{lemma}[Least upper bound]
\label{thm:lub}
For ground $\sigma_1$ and $\sigma_2$, if $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ then $\sigma$ is a least upper bound by $\sqsupseteq$.
\end{lemma}

\begin{figure}[!h]
\vspace{0em}  
\noindent
\begin{equation*}
\begin{array}{rclcrclcrcl}
 \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
 \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
\end{array}
\end{equation*}
\noindent
\vspace{-0.5em}
\begin{equation*}
\begin{array}{l}
 \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
 \\[0.5em]
 \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}_{\rho_i}} =\\[0.1em]
 \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n}, \addopt{\theta(\rho_i)} \}
 \\[0.5em]
 \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
 \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i \\[0.6em]
 \textnormal{Choose minimal}\ \theta\ \textnormal{by ordering}\ \sqsubseteq\ \textnormal{lifted over substitutions} \\[0.1em]
\end{array}
\end{equation*}
\caption{Shape inference from sample data}
\label{fig:shape-inference}
\vspace{-0.5em}
\end{figure}

\begin{proof}
By induction over the definition of $\ident{csh}$. Note that the algorithm only infers the top shape 
\kvd{any} when on of the shapes is the top shape (\emph{top}) or when there is no other option 
(\emph{any}); a nullable shape is introduced in $\addopt{-}$ only when no non-nullable shape can
be used (\emph{null}), (\emph{opt}).
\end{proof}

@@

Figure~\ref{fig:shape-inference} defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a 
collection of sample data $d_1, \ldots, d_n$ into a shape $\sigma$. Before applying $S$, we assume 
each record in each $d_i$ is marked with a fresh row inference variable $\rho_i$.

We then choose a ground, minimal, global substitution $\theta$ for row variables. The substitution
is minimal in the sense that it introduces as few new fields as possible, but no additional fields, 
and the newly introduced fields have the most specific shape. Because $\rho_i$ variables represent 
potentially missing fields, the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} is applied 
to all types in the vector.

This is sufficient to equate the record field labels and satisfy the pre-conditions in rule 
(\emph{recd}) when multiple record samples are merged. In practice, $\theta$ is found via row 
variable unification \cite{rows-remy}. We omit the details here. No $\rho_i$ variables remain after 
inference as the substitution chosen is ground.

@@

\begin{figure*}
  \noindent
  \begin{equation*}
  \begin{array}{rclcl}
   \tytag &\narrow{=}& \ident{collection}  &\narrow{\lsep}& \ident{number} \\
          &\narrow{\lsep}& \ident{nullable} &\narrow{\lsep}& \ident{string}  \\
          &\narrow{\lsep}& \nu ~\lsep \ident{any} &\narrow{\lsep}& \ident{bool}
  \end{array}
  %
  \quad\;\;\;
  %
  \begin{array}{rcl}
   \tytagof(\ident{string}) &\narrow{=}& \ident{string}\\
   \tytagof(\ident{bool}) &\narrow{=}& \ident{bool}\\
   \tytagof(\ident{int}) &\narrow{=}& \ident{number}\\
   \tytagof(\ident{float}) &\narrow{=}& \ident{number}\\
  \end{array}
  %
  \;\;
  %
  \begin{array}{rcl}
   \tytagof(\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) &\narrow{=}& \ident{any}\\
   \tytagof(\nu\; \{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \nu \\
   \tytagof(\kvd{nullable}\langl\hat{\sigma}\rangl) &\narrow{=}& \ident{nullable}\\
   \tytagof([\sigma]) &\narrow{=}& \ident{collection}\\
  \end{array}
  \end{equation*}

  \vspace{-0.5em}

  \begin{equation*}
  \qquad\qquad
  \begin{array}{ll}
  \begin{array}{l}
    \ident{csh}(\kvd{any}\langl \sigma_1, \ldots, \sigma_k,  \ldots, \sigma_n\rangl, 
      \kvd{any}\langl \sigma'_1, \ldots, \sigma'_k, \ldots, \sigma'_m\rangl) =\\
      \qquad \kvd{any}\langl \ident{csh}(\sigma_1, \sigma_1'), \ldots, \ident{csh}(\sigma_k, \sigma_k'), 
        \sigma_{k+1}, \ldots, \sigma_{n}, \sigma'_{k+1}, \ldots, \sigma'_{m}\rangl \\[0.5em]
      \qquad \textnormal{For $i,j$ such that }(\tytagof(\sigma_i) = \tytagof(\sigma'_j)) \Leftrightarrow (i = j) \wedge (i \leq k)
  \end{array}\quad~
  &  (\emph{top-merge}) 
  \\[2.5em]    
  \begin{array}{l}
    \ident{csh}(\sigma, \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) = 
    \ident{csh}(\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl, \sigma) = \\
    \qquad \kvd{any}\langl\sigma_1, \ldots, \dropopt{\ident{csh}(\sigma, \sigma_i)}, \ldots, \sigma_n\rangl\\[0.5em]
    \qquad \textnormal{For $i$ such that }\tytagof(\sigma_i) = \tytagof(\dropopt{\sigma})
  \end{array}
  &  (\emph{top-incl}) 
  \\[2.5em] 
  ~\;\ident{csh}(\sigma, \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) =
   \kvd{any}\langl\sigma_1, \ldots, \sigma_n, \dropopt{\sigma}\rangl
   & (\emph{top-add})\\[0.5em]    
  ~\;\ident{csh}(\sigma_1,\sigma_2) =
  \kvd{any}\langle\dropopt{\sigma_1}, \dropopt{\sigma_2}\rangle
  & (\emph{top-any})
 \end{array}
\end{equation*}

\caption{Extending the common preferred shape relation for labelled top shapes}
\label{fig:subtyping-cst-var}
\end{figure*}

@@

\section{Relative type safety}
\label{sec:safety}
